---
title: "Spotify Track Breakdown : IST 719 Wed 8pm Section "
author: "Jill Davis"
date: "08/18/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data

```{r load in the data}

library(readxl)
spotify_data <- read.csv( "/Users/jill/Downloads/archive/tracks.csv"
                          , header = TRUE
                          , stringsAsFactors = FALSE)
```

# Section 3
```{r decades}
# First, group data by decade by adding a column 
library(dplyr)
spotify_data <-
        spotify_data%>%
 mutate(decade = case_when(
         startsWith(release_date, "192") ~ "1920s"
 ))

# Then, add subsequent decades by subsetting the data 
spotify_data$decade[spotify_data$release_date < 1940 & spotify_data$release_date >= 1930] <- c("1930s")
spotify_data$decade[spotify_data$release_date < 1950 & spotify_data$release_date >= 1940] <- c("1940s")
spotify_data$decade[spotify_data$release_date < 1960 & spotify_data$release_date >= 1950] <- c("1950s")
spotify_data$decade[spotify_data$release_date < 1970 & spotify_data$release_date >= 1960] <- c("1960s")
spotify_data$decade[spotify_data$release_date < 1980 & spotify_data$release_date >= 1970] <- c("1970s")
spotify_data$decade[spotify_data$release_date < 1990 & spotify_data$release_date >= 1980] <- c("1980s")
spotify_data$decade[spotify_data$release_date < 2000 & spotify_data$release_date >= 1990] <- c("1990s")
spotify_data$decade[spotify_data$release_date < 2010 & spotify_data$release_date >= 2000] <- c("2000s")
spotify_data$decade[spotify_data$release_date < 2020 & spotify_data$release_date >= 2010] <- c("2010s")
spotify_data$decade[spotify_data$release_date < 2030 & spotify_data$release_date >= 2020] <- c("2020s")

# verifying changes:
head(spotify_data)
tail(spotify_data)

# Looks good! 
```

# Time to Plot

```{r track duration}
barplot(tapply(spotify_data$duration_ms, spotify_data$decade, mean)
        , main = "Track duration over the decades"
        , ylab = "Average Track Duration (in milliseconds)"
        , col = c("#FBA164", "#0C5061", "#D6E0E4")) # pretty aqua color #50A3A4
```

# Does track duration correlate with track popularity?

```{r duration v popularity}
plot(spotify_data$duration_ms/60000, spotify_data$popularity
     , pch = 16
     , main = "track popularity by track duration"
     , xlab = "duration (minutes)"
     , ylab = "popularity"
     , col = "#0f2a56"
     , bty = "n")

# It appears shorter songs are more popular in this data set
```

# Danceability (for funsies)

```{r track danceability over time}
plot(tapply(spotify_data$danceability, spotify_data$decade, mean)
        , main = "Average Danceability Ratings from 1920 through 2020"
        , ylab = "Danceability"
        ,xlab = "Decades"
        , col = c("#0C5061")
        , type = "l"
        , bty = "n")
# Danceability seems to remain about the same over time
mean(spotify_data$danceability)
# = .56, about the same very average
```

# Is there a set of track traits which embody the sound of a specific time period?

```{r zeitgeist}
# not sure how to do this
# attempts in ggplot: 
# Attempt 1 : Managed to get the points in different colors, missing a line and a legend
a <- tapply(spotify_data$loudness, spotify_data$decade, mean)
b <- tapply(spotify_data$speechiness, spotify_data$decade, mean)
c <- tapply(spotify_data$acousticness, spotify_data$decade, mean)
d <- tapply(spotify_data$instrumentalness, spotify_data$decade, mean)
e <- tapply(spotify_data$liveness, spotify_data$decade, mean)
spotify_data_zeitgeist <- data.frame(a,b,c,d,e)

colnames(spotify_data_zeitgeist)
renamed <- c("mean.loudness", "mean.speechiness", "mean.acousticness"
             , "mean.instrumentalness", "mean.liveness")
colnames(spotify_data_zeitgeist) <- renamed

spotify_data_zeitgeist$decade <- c("1920s", "1930s", "1940s", "1950s", "1960s"
                                   , "1970s", "1980s", "1990s", "2000s", "2010s"
                                   , "2020s")
library(ggplot2)
mean_plot <- ggplot(spotify_data_zeitgeist) + aes(x = decade) + #group = 1?
        #geom_point(aes(y = mean.loudness)) +
        geom_point(aes(y = mean.speechiness), col = "red") +
        geom_point(aes(y = mean.acousticness), col = "blue") +
        geom_point(aes(y = mean.instrumentalness), col = "purple") +
        geom_point(aes(y = mean.liveness), col = "green") 
mean_plot <- mean_plot + labs(y = "", x = "") + ggtitle("Average Rating for Each Decade")
mean_plot

# multi=series bar chart (?)


```

# Customizing the sounds of the decades plot

```{r stacked bar chart for average songs}
# first, convert the data frame to a matrix to pass through barplot
spotify_data_zeitgeist_matrix <- as.matrix(spotify_data_zeitgeist)
renamed_4legend <- c("loudness", "speechiness", "acousticness"
             , "instrumentalness", "liveness", "decade")
colnames(spotify_data_zeitgeist_matrix) <- renamed_4legend
# then, transpose the data because the first time I did this it was sideways
spotify_data_zeitgeist_matrix <- t(spotify_data_zeitgeist_matrix)
# get rid of mean.loudness because it is an outlier and hides the rest of the plot
spotify_data_zeitgeist_matrix <- spotify_data_zeitgeist_matrix[-1,]
spotify_data_zeitgeist_matrix <- spotify_data_zeitgeist_matrix[-5,]

barplot(spotify_data_zeitgeist_matrix, border = NA, space = 0.25,
        xlab = "", ylab = "rating", main = "sounds of the decades"
        , sub = "mean value of audio track characteristics for the last one-hundred years"
        , legend.text = TRUE
        , col = c("#0f2a56","#2a8897","#e8a266", "#e66b37"))
```

# Customizing the duration plot

```{r scatter plot with high density lines}
# Then, plot the points
# I want to change milliseconds to minutes, divide vector by 60,000, but only once:
spotify_data$duration_ms <- (spotify_data$duration_ms)/60000
plot((tapply(spotify_data$duration_ms, spotify_data$decade, mean))
      , type="p", bty = "n") # wanted to do y-axis in descending order, making a vector of a range:
range.duration <- c(0,5)
# combining high density vertical lines (h) with p
plot((tapply(spotify_data$duration_ms, spotify_data$decade, mean))
     , type="h", ylab = "Duration (in minutes)", xlab = "Decade"
     , main = "Average Lengths of Audio Tracks, 1920-2020", ylim = range(range.duration)
     , bty = "n")
points((tapply(spotify_data$duration_ms, spotify_data$decade, mean))
       , pch=19, col="#0047ab", cex = 1.4, bty = "n")
```

# Trying to flip the duration chart so it looks like music
```{r rise from the ashes}
# I want the years to show in descending order:
barplot((tapply(spotify_data$duration_ms, spotify_data$decade, mean))
        , horiz = TRUE)
barplot(rev(tapply(spotify_data$duration_ms, spotify_data$decade, mean))
        , horiz = T
        , las = 1
        , xlab = "duration (in minutes)"
        , main = "average length of audio tracks, 1920-2020"
        , las = 1
        , col = c("#0f2a56")
        , space = 2
        , bty = "n"
        #, xlim = range.duration
        )
```

# Since the network didn't work out, decided to do a wordcloud of song titles:

```{r seventies titles}
# wordcloud steps from :
# https://towardsdatascience.com/create-a-word-cloud-with-r-bde3e7422e8a 
# load libraries
library(wordcloud)
library(RColorBrewer)
library(tm)

# tried to take remaster / remastered out but failed
spotify_data$name <- gsub("remaster", "", spotify_data$name)
spotify_data$name <- gsub("remastered", "", spotify_data$name)

text <- spotify_data$name[spotify_data$decade == "1970s"]
titles <- Corpus(VectorSource(text))

# Clean the data
titles <- titles %>%
  tm_map(removeNumbers) %>%
  tm_map(removePunctuation) %>%
  tm_map(stripWhitespace)
titles <- tm_map(titles, content_transformer(tolower))
titles <- tm_map(titles, removeWords, stopwords("english"))

# Create a data frame that contains each word and their frequency:
dtm <- TermDocumentMatrix(titles) 
matrix <- as.matrix(dtm) # error: vector memory exhausted, even after upping limit
words <- sort(rowSums(matrix),decreasing=TRUE) 
df <- data.frame(word = names(words),freq=words)

#generate the word cloud:
set.seed(1234) # for reproducibility 
wordcloud(words = df$word, freq = df$freq, min.freq = 1, max.words=200
          , random.order=FALSE, rot.per=0.35
          , colors = c("#0f2a56","#2a8897","#e8a266", "#e66b37"))
```
# Finally, I need some single dimension plots showing distribution ...

```{r tempo}
hist(spotify_data$tempo
     , border = NULL
     , xlab = "tempo"
     , ylab = "frequency"
     , col = "#0f2a56"
     , main = "tempo distribution")
```
# All data from : https://www.kaggle.com/search?q=spotify+1922-2021+in%3Adatasets 
